---
title: "Assignment 2 Script"
author: "Christina Treacy"
date: "May 7, 2026"
format: pdf
editor: visual
---

```{r}
install.packages("dataRetrieval")
library(dataRetrieval)
library(tidyverse)
library(here)
```

Load in data from NWIS server and save to CSV

```{r}
siteNumbers <- c("14211720", "14211010", "14166000")
parameterCd <- "00010" # water temperature
startDate <- "2010-01-01"
endDate <- "2023-12-31"
temp_data <- readNWISdv(siteNumbers, parameterCd, startDate, endDate)

# rename columns for clarity
colnames(temp_data) <- c("agency_cd", "site_no", "Date", "temp_C")
# extract some date info for use later
temp_data$month = month(temp_data$Date, label=TRUE)
temp_data$year = year(temp_data$Date)

write.csv(temp_data, "tempdata.csv", row.names = FALSE)
```

Load data directly from repository

```{r}
tempdata <- read.csv(here("tempdata.csv"))
```

Plot data & Calculate Summary Statistics

```{r}
# check by plotting
ggplot(tempdata, aes(x=Date, y=temp_C, color=site_no)) +
geom_line() +
labs(title="Stream Temperature at Willamette River Sites",
x="Date", y="Temperature (°C)", color="Site Number") +
theme_minimal()

# generate summary stats
summary_table <- tempdata |>
  group_by(site_no) |>
  summarize(min=min(temp_C), 
            mean=mean(temp_C),
            med=median(temp_C),
            max=max(temp_C),
            st_dev=sd(temp_C)) 
write.csv(summary_table, here("site_temp_summary_table.csv"))
```

Trends in Water Temperature for the month of February
```{r}
# filter for February temperatures
feb_data <- subset(temp_data, month == "Feb")

# fit a GLM model to see if there is a trend in Feburary temperatures over time, separated by site
feb_model <- glm(temp_C ~ year*site_no, data = feb_data)
summary(feb_model)

```

The slope for first site is -0.064277, showing a slight negative trend in water temperature over time in the month of February.

Site 14211010 has a slope that is 0.019204 greater than the first site, and Site 14211720 has a slope that is 0.01619 greater than the first site. Both of these results demonstrate relatively small differences in trends between sites.However, the p-value is
greater than 0.05 for both of these estimates, meaning that these results are not statistically significant.
